---
title: "BCB420 Assignment 1- Data set selection and initial processing"
author: "Angela Ng"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

# Introduction

# Download the data

Firstly, we need to download the dataset containing the counts from GEO.  We 
want to do it so that it is only downloaded once initially.

```{r, message=FALSE, warning=FALSE}
# Install required packages 
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
if (!requireNamespace("GEOquery", quietly = TRUE)) {
  BiocManager::install("GEOquery")
}

gse <- "GSE116250"
gseDirectory <- "/home/rstudio/projects/GSE116250"
if (!file.exists(gseDirectory)) {
  # Get supplementary files from GEO
  suppFiles <- GEOquery::getGEOSuppFiles(gse)
  fileName <- rownames(suppFiles)[[1]]
}
```
There is only one file from the data set downloaded from GEO.  This is the 
`GSE116250_rpkm.txt.gz`.  We can now read in the file

```{r, tidy=TRUE}
expData <- read.delim(fileName, header = TRUE, check.names = FALSE)
colnames(expData)
dim(expData)
```
From this we see there are 57974 rows and 66 columns in this data set.

# Compute statistics to assess data quality for control and test conditions in dataset

# Map rows to symbols

# Clean

# Questions to address

1. What are the control and test conditions of the dataset?
2. Why is the dataset of interest to you?
3. Were there expression values that were not unique for specific genes? How did you handle these?
4. Were there expression values that could not be mapped to current HUGO symbols?
5. How many outliers were removed?
6. How did you handle replicates?
7. What is the final coverage of your dataset?

# References
